---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sfdep

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/sfdep)](https://CRAN.R-project.org/package=sfdep)
<!-- badges: end -->

sfdep builds on the great shoulders of spdep package for spatial dependence. sfdep creates an sf and tidyverse friendly interface to the package as well as introduces new functionality that is not present in spdep. sfdep utilizes list columns extensively to make this interface possible. 

## Installation

You can install the development version of sfdep like so:

``` r
remotes::install_github("josiahparry/sfdep")
```

## Usage

There are three main categories of functionality relating to geometry neighbors, weights, and local indicators of spatial association (LISAs). 

### Neighbors

The most fundamental usage is to find contiguous neighbors from a polygon. This is done with `st_contiguity()` which, by default creates queen weights. If rook weights are desired, set `queen = FALSE`. Additional arguments can be passed to the underlying `spdep::poly2nb()` via `...`. `st_contiguity()` creates an object of class `nb` as used by `spdep`. 

```{r message=FALSE}
library(sf)
library(sfdep)
library(dplyr)

# grab geometry
geo <- st_geometry(guerry)

nb <- st_contiguity(geo)

nb
```

We can identify higher order neighbors with `st_nb_lag()` and the cumulative higher order neighbors with `st_nb_lag_cumul()`.

```{r}
st_nb_lag(nb, 2)
st_nb_lag_cumul(nb, 2)
```

Other point geometry neighbor functions are `st_knn()`, `st_nb_band()`, `st_nb_dists()`. 


### Weights


Polygon weights are created with `st_weights()` (which calls `spdep::nb2listw`). By default they are row standardized weights. 

```{r}
wt <- st_weights(nb)

wt[1:2]
```

Other point based weights can be created with `st_kernel_weights()` and `st_inverse_weights()`.


### Local Indicators of Spatial Association (LISAs)

LISAs are created from a combination of neighbors and weights and are intended to be used inside of a dplyr pipeline. The below is a worked example of calculating the spatial lag and the local moran.

```{r, message = FALSE, warn = FALSE}
g <- guerry %>% 
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb))
```

Then calculate the spatial lag with `st_lag()`. Given that we've only modified an sf object, we can visualize this with ggplot2.

```{r}
library(ggplot2)

# create spatial lag
g %>% 
  mutate(crime_pers_lag = st_lag(crime_pers, nb, wt)) %>% 
  ggplot(aes(fill = crime_pers_lag)) + 
  geom_sf(lwd = 0.2, color = "black") +
  theme_void()
```

Most users will be interested in local indicators of spatial association (LISA). Utilize `local_moran()` to do this. `local_moran()` will create a data frame column which contains a number of informative variables. For example the cluster that a polygon falls into based on mean, median, or pysal calulations. This will need to be unnested or certain variables hoisted.

Create the Local Moran data frame column.
```{r}
lisa <- g %>% 
  mutate(moran = local_moran(crime_pers, nb, wt))

pull(lisa, moran) %>% 
  glimpse()
```

Visualize this.

```{r}
lisa %>% 
  tidyr::unnest(moran) %>% 
  ggplot(aes(fill = mean)) +
  geom_sf() +
  geom_sf(lwd = 0.2, color = "black") +
  theme_void() +
  scale_fill_manual(values = c("#1C4769", "#24975E", "#EACA97", "#B20016"))
```

<!-- ## Neighbor apply function -->

<!-- In the case that you want to create custom local level metrics, you can do so with the function `st_nb_apply()`. This provides a purrr like interface to do calculation on an observation $x_{i}$ using it's neighbors, $x_{ij}$, and their respective weights $w_{ij}$. In the formula syntax utilize `.xij`, `.nb` and `.wt` for these values respectively. We can specify the type of output with the suffix argument. By default this will be `"dbl"`. These can be any purrr sufix. To get the default list output use `"list"`.  -->

<!-- For example  -->

<!-- ```{r} -->
<!-- g %>%  -->
<!--   transmute(x = st_nb_apply(crime_pers, nb, wt,  -->
<!--                             .f = function(.xij, .wt, ...) sum((.xij - crime_pers)^2) / (length(.wt) + 1), suffix = "dbl")) -->
<!-- ``` -->


## Development Tracking

### Pysal ESDA functionality

#### Global Statistics

- [] Moran's I
- [] Join Count
- [] Gettis-Ord

#### LISAs

- [x] Local Moran
  - [] Bivariate local moran
- [] Local Join Count
  - [x] Univariate
  - [] Bivariate
    - No colocation
    - colocation
  - [] Multivarite
- [x] LOSH
- [x] Gi 
- [x] Gi*

### Utilities

- [x] Conditional permutation
