---
title: "rstudio::conf(2022L)"
subtitle: Exploratory Spatial Data Analysis in the tidyverse
output: html_notebook
---

* Hello it me


* I studied Geographic Information Systems in college where I was taught how to open menus and click button in ESRI's toolsuite. Discovered R and thus other open-source tools like QGIS and python. Became an OSS convert. 
* During college, I taught myselfg enough to be dangerous with the tidyverse. 
  * I wanted to do my spatial analysis work in R. I couldn't get my head around sp and s4 classes. I gave up and continued to work in QGIS
* 2018 rolls around and two things happen: I went to graduate school and the previously naiscent R package {sf} had reached full maturity with stable support and integration with dplyr and ggplot2
* Impact of {sf}:
  * {sf} is probably the most revolutionary spatial R package for programmers of my generation
  * the ability to couple familiar workflows using dplyr and ggplot2 cannot be understated. 
  * seamlessly go from analysis, to visualization, and basic geospatial operations
* 2019 took Advanced Spatial Analysis where I was taught pysal, shapely, and geopandas
  * its not R.
  * R needs a spatial analysis library to compete 
* discovered {spdep}, a great library, but again, idiosyncratic, and most importantly to me, not compatible with the tidyverse meaning it cannot be so easily adopted
* spdep is a package for spatial weights and statistics
  * it is somewhat of a legacy package, 
    * it's first release was in 2002 (for context I was in the 1st grade)
* inspired {sfdep} - is a tidy interface to spdep and extends it by providing additional statistics and operations
* sfdep design principles
* comparison of syntax

```{r}
library(spdep)
library(sfdep)
library(dplyr)
library(ggplot2)


library(spdep)
guerry <- sfdep::guerry
nb <- poly2nb(guerry)
listw <- nb2listw(nb)
lm_sp <- localmoran_perm(guerry$crime_pers, listw, nsim = 199)
quads <- attr(lm_sp, "quadr")
res <- cbind(guerry, lm_sp, quads)



# note the attributes store the quadrant  which we want for visualization

res |> 
  mutate(`Local Moran` = ifelse(Pr.folded..Sim <= 0.1, Ii, NA)) |> 
  ggplot(aes(fill = `Local Moran`)) +
  geom_sf(lwd = 0.1, color = "grey60") + 
  scale_fill_viridis_c(option = "A", na.value = "white") +
  theme_void() +
  labs(title = "Autocorrelation of Crime", 
       subtitle = "France, 1833") +
  theme(title = element_text(face = "bold")) 



# Using sfdep
# can all be done in one pipe
library(sfdep)
res <- guerry |> 
  dplyr::mutate(nb = st_contiguity(geometry),
                wt = st_weights(nb),
                lm = local_moran(crime_pers, nb, wt)) |> 
  tidyr::unnest(lm) 
head(res)

# res |> 
#   mutate(`Local Moran` = ifelse(p_folded_sim <= 0.1, ii, NA)) |> 
#   ggplot(aes(fill = `Local Moran`)) +
#   geom_sf(lwd = 0.1, color = "grey60") + 
#   scale_fill_viridis_c(option = "A", na.value = "white") +
#   theme_void() +
#   labs(title = "Autocorrelation of Crime", 
#        subtitle = "France, 1833") +
#   theme(title = element_text(face = "bold")) 
# 

library(sfdep)
library(ggplot2)
res <- guerry |> 
  dplyr::mutate(nb = st_contiguity(geometry),
                wt = st_weights(nb),
                lm = local_moran(crime_pers, nb, wt)) |> 
  tidyr::unnest(lm) 

res |> 
  dplyr::mutate(Cluster = ifelse(p_folded_sim <= 0.1, as.character(pysal), NA)) |> 
  ggplot(aes(fill = Cluster)) +
  geom_sf(lwd = 0.1, color = "black") + 
  scale_fill_manual(
    values = c("High-High" = "#bd4f6b",
               "High-Low" = "#db6381",
               "Low-Low" = "#5084a3",
               "Low-High" = "#8cc5e6"),
    na.value = "#fcfcf2"
    ) + 
  theme_void() +
  labs(title = "Clusters of Crime", 
       subtitle = "France, 1833") +
  theme(title = element_text(face = "bold")) 
```


* generating neighbors:
  * contiguity
  * knn
  * distance
* 


```{r}
library(sfdep)
df_fp <- system.file("extdata", "bos-ecometric.csv", package = "sfdep")
geo_fp <- system.file("extdata", "bos-ecometric.geojson", package = "sfdep")

# read in data
df <- readr::read_csv(df_fp, col_types = "ccidD")
geo <- sf::read_sf(geo_fp)

# create spacetime object
bos <- spacetime(df, geo, ".region_id", "year")

ehsa <- emerging_hotspot_analysis(bos, "value", include_gi = TRUE)

ehsa_sf <- left_join(geo, ehsa, by = c(".region_id" = "location"))

ggplot() +
  geom_sf(
    aes(fill = classification),
    ehsa_sf, lwd = 0.05, color = "black"
    ) +
  scale_fill_manual(
    values = c(
      "no pattern detected" = "#fffff2", 
      "persistent coldspot" = "#6ba5c7",
      "sporadic coldspot" = "#5084a3",
      "sporadic hotspot" = "#db6381", 
      "oscilating hotspot" = "#bd4f6b",
      "consecutive hotspot" = "#a3485f",
      "persistent hotspot" = "#732c3e")
    ) +
  theme_light() +
  labs(
    title = "Prevalence of Guns in Suffolk County",
    subtitle = "Emerging Hot Spot Analysis",
    caption = "Missing data is represented by an empty polygon."
  )


```


```{r}
bos_sf <- bind_cols(df, attr(ehsa, "gi_star")) |> 
  left_join(geo) |> 
  sf::st_as_sf()

ggplot(bos_sf, aes(fill = gi_star)) +
  geom_sf(color = NA) +
  facet_wrap("year", ncol = 4) +
  scale_fill_gradient2(
    high = scales::muted("red"),
    low = scales::muted("blue")
    ) + 
  labs(title = "Space-time-lagged Gi*",
       subtitle = "Prevelance of Guns from 2010 - 2019") + 
  theme_void() 
```

```{r}
library(sfdep)
library(dplyr)

g_ntwk <- guerry |> 
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb)) |> 
  st_as_graph()

plot(g_ntwk)
```



* What is Exploratory Data Analysis?
* Extending EDA to spatial data is ESDA
* Spatial data: point, lines, polygons, and pictures
  * Show points (Crime locations)
  * lines (road networks)
  * polygons (census regions)
* Tobler's First Law
* Questions that ESDA can help answer:
  * Are things randomly distributed? 
  * Are there spatial outliers?
* ESDA compares values to neighbors as opposed to the whole
