---
title: "WIP: Understanding Emerging Hot Spots"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



Emerging hot spot Analysis (EHSA) is a technique that falls under exploratory spatial data analysis (ESDA). It combines the traditional ESDA technique of hot spot analysis using the Getis-Ord Gi* statistic with the traditional time-series Mann-Kendall test for monotonic trends. 

The goal of EHSA is to evaluate how hot and cold spots are changing over time. It helps us answer the questions: are they becoming increasingly hotter, Are they cooling down, or are they staying the same? 

In brief, EHSA works by calculating the Gi* for each time period. The Gi* at each location is treated as a time-series and evaluated for a trend using the Mann-Kendall statistic. In this vignette we walk through the Gettis-Ord Gi\*, the Mann-Kendall, and how the two work together to in EHSA. The Gi\* and the Mann-Kendall are compared together to create 17 unique classifications to help better understand how the locations have changed over time. 

## Gettis-Ord Gi*

The Gettis-Ord Gi and Gi* (pronounced gee-eye-star) are one of the earliest LISAs.
The Gi and Gi* measures are typically reported as a Z-score where high values indicate
a high-high cluster and negative Z-scores indicate a low-low cluster. There are no 
high-low and low-high classifications like the local Moran.

"The Gi statistic consist of a ratio of the weighted average of the values in the neighboring locations, to the sum of all values, not including the value at the location ($x_i$)" ([Local Spatial Autocorrelation (2), GeoDa Center](https://geodacenter.github.io/workbook/6b_local_adv/lab6b.html#getis-ord-statistics)). 

$$
G_i = \frac{\sum_{j \ne i}W_{ij}X_j}
          {\sum_{j \ne i}X_j}
$$

The Gi* statistic includes the focal (or self, or ith) observation in the neighborhood.

$$
G_i* = \frac{\sum_{j}W_{ij}X_j}
          {\sum_{j}X_j}
$$

### Implementing the local Gi*

To calculate the local Gi* using sfdep, we have to be especially aware of the neighbors
list when we create it. By default, when we create a neighbors list, we exclude the
self from the observation. However, if we want to calculate the local Gi*, we must 
be sure to explicitly add it using `include_self()`

Here, we follow the example laid out by the GeoDa center documentation to calculate and plot the local Gi* statistic for donations using the Guerry dataset. 

Here we create a neighbor list and include self and create weights from the neighbors. 

```{r message = FALSE}
library(sfdep)
library(dplyr)

guerry_nb <- guerry |> 
  mutate(nb = include_self(st_contiguity(geometry)),
         wt = st_weights(nb)
  ) 
```

Following, we calculate the local Gi* using `local_gstar_perm()` on the `donations` column which creates a new data frame column called `gistar`. We then unnest it using `tidyr::unnest()`

```{r}
donat_gistar <- guerry_nb |> 
  transmute(gistar = local_gstar_perm(donations, nb, wt, nsim = 999)) |> 
  tidyr::unnest(gistar)

donat_gistar
```

Lastly, we classify the clusters using a combination of `mutate()` and `case_when()` 
which is then piped into a ggplot map. While not a perfect recreation of the GeoDa map, it is very close—likely due to conditional permutation (see [conditional permutation vignette](/articles/conditional-permutation.html) for more on significance calculation).

```{r}
library(ggplot2)

donat_gistar |> 
  mutate(cluster = case_when(
    p_folded_sim > 0.05 ~ "Not Significant",
    p_folded_sim <= 0.05 & gi_star < 0 ~ "Low",
    p_folded_sim <= 0.05 & gi_star > 0 ~ "High"
  )) |> 
  ggplot(aes(fill = cluster)) +
  geom_sf(lwd = 0.2, color = "black") +
  scale_fill_manual(values = c("High" = "red",
                               "Low" = "Blue", 
                               "Not Significant" = "white")) +
  theme_void()
```

In EHSA, we calculate this statistic for each location and every time period in our dataset. 

## Mann-Kendall Test

- tests if there is a monotonic trend either in a positive or negative direction
- if you dont remember what monotonic means, a monotonic series is one that is always increasing and never decreasing or vice versa. 
- so a monotonic trend consistently increases (or decreases) through time. And, importantly the trend doesn't have to be linear. 

Below is an example of a monotonic upward trend which isnt linear

```{r}
series <- c(0, 1, 1.4, 1.5, 1.6, 5, 5, 5.5, 20, seq(20, 21, length.out = 5))

plot(series, type = "l")
```


- null hypothesis: no trend

https://www.real-statistics.com/time-series-analysis/time-series-miscellaneous/mann-kendall-test/



## Emerging Hot Spot Analysis by Hand

```{r setup}
library(sfdep)
library(dplyr)


# replicate the guerry dataset 10 times
x <- purrr::map_dfr(1:10, ~guerry) |> 
  select(code_dept, crime_pers) |> 
         # create an indicator for time period
  mutate(time_period = sort(rep(1:10, 85)), 
         # add some noise 
         crime_pers = crime_pers * runif(850, max = 2))

spt <- as_spacetime(x, .loc_col = "code_dept", "time_period")
```

Here we calculate emerging hot spots manually. However, here we cannot include time lag because there's no outside of the box functionality to do so.


This chunk:

- activates the geometry context and:
  - creates neighbors
  - weights
- activates the data context
- sets the neighbor column from the geometry context
- sets the weights column from the geometry context
- groups by `time_period`
- calculates the local Gi* for each time period

```{r}
gistars <- spt |> 
  activate("geometry") |> 
    mutate(nb = include_self(st_contiguity(geometry)),
         wt = st_weights(nb)) |> 
  activate("data") |> 
  set_nbs("nb") |> 
  set_wts("wt") |> 
  group_by(time_period) |> 
  mutate(gi_star = local_gstar_perm(crime_pers, nb, wt))

```

here we see the Gi* for one location for all time periods. 

```{r}

dept_01_trend <- gistars |> 
  tidyr::unnest(gi_star) |> 
  filter(code_dept == "01") |> 
  select(gi_star, time_period, p_sim)

dept_01_trend
```

Then we calculate the Mann-kendall statistic for it

```{r}
Kendall::MannKendall(dept_01_trend[["gi_star"]])
```

Here we have an insignificant positive trend

next we 


```{r}
x |> 
  group_by(time_period) |> 
  arrange(time_period, code_dept) |> 
  mutate(nb = include_self(st_contiguity(geometry)),
         wt = st_weights(nb),
         gi_star = local_gstar(crime_pers, nb, wt)) |> 
  group_by(code_dept) |> 
  summarise(trend = broom::tidy(Kendall::MannKendall(gi_star))) |> 
  tidyr::unnest(trend) |> 
  arrange(p.value)
```
Here we use emerging hot spot analysis function from sfdep using no time lag

```{r}
spt <- x |> 
  as_spacetime("code_dept", "time_period")

emerging_hotspot_analysis(spt, "crime_pers", k = 0) |> 
  arrange(p_value)
```


#### Sources

- GeoDa center: https://geodacenter.github.io/workbook/6b_local_adv/lab6b.html#getis-ord-statistics



