---
title: "Version 0.2.0"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sfdep)
```


While not too far from the initial CRAN release of sfdep, this version consists of two sizeable additions relating to spatio-temporal analysis and colocation analysis as well as bug fixes and improvements. Additional functionality is added regarding point-pattern analysis and neighbor / weight list utilities. Given the breadth of the additions, this represents a minor version release. 


## Colocation Analysis

This release includes functionality to compute colocation quotients. This is extremely exciting as this is the first open source implementation of Colocation Quotients that I am aware of. There are three types of colocation quotient (CLQ) in this release: the global CLQ, pairwise CLQ, and local CLQ.

See the [colocation vignette](/articles/colocation-analysis.html) for a more detailed write up. 

There are three new functions for calculating CLQs:

- `global_colocation()`
- `pairwise_colocation()`
- `local_colocation()`


## Spacetime 

And probably most notable, is the introduction of a new spacetime class. This class was created as a byproduct of creating functionality for emerging hot spot analysis. For a more detailed write up see the [spacetime vignette](/articles/spacetime-s3.html).

The new functions that are available are:

- `spacetime()`, `new_spacetime()`, `validate_spacetime()`:
  - Creating spacetime objects
- `activate()`, `active()`:
  - changing spacetime context and determining active context
- `as_sf()`, `as_spacetime()`:
  - casting between spacetime and sf objects
- `is_spacetime()`: 
  - determine if an object is a spacetime object
- `is_spacetime_cube()`:
   - determine if a spacetime object is a spatio-temporal full grid aka spacetime cube
- `complete_spacetime_cube()`:
  - if a spatio-temporal sparse grid, add rows to make a spatio-temporal full grid (spacetime cube)
- `set_col()`, `set_nbs()`, `set_wts()`:
  - add columns from geometry context to data context
- `emerging_hotspot_analysis()`:
  - conduct emerging hotspot analysis
- `spt_update()`:
  - update times and locations attributes of a spacetime object
  
```{r}
df_fp <- system.file("extdata", "bos-ecometric.csv", package = "sfdep")
geo_fp <- system.file("extdata", "bos-ecometric.geojson", package = "sfdep")

# read in data
df <- readr::read_csv(df_fp, col_types = "ccidD")
geo <- sf::read_sf(geo_fp)

# create spacetime object
bos <- spacetime(df, geo, ".region_id", "year")

bos
```

Spacetime objects have two contexts: data and geometry. Currently the data is activated. We can activate the geometry context like so:

```{r}
activate(bos, "geometry") 
```
This is handy because we can find neighbors in our geometry and link them to our data.

```{r}
library(dplyr)

bos_nb <- bos |> 
  activate("geometry") |> 
  mutate(nb = st_contiguity(geometry),
         wt = st_weights(nb))

bos_nb
```
These can be brought over to our data context for further use.

```{r}
bos_nb <- bos_nb |> 
  set_wts() |> 
  set_nbs()

bos_nb
```

Then we can group our dataset and calculate metrics based on different years. But this is only possible for data that are spacetime cubes. Read the spacetime vignette for more. We can check if this object meets the conditions to be a spacetime cube.
  
```{r}
is_spacetime_cube(bos)
```

Since this is a spacetime cube, it is safe to perform statistics on each time slice.

```{r}
bos_gs <- bos_nb |> 
  activate("data") |> 
  filter(ecometric == "Guns") |> 
  group_by(year) |> 
  mutate(g = local_g(value, nb, wt))

bos_gs
```
Now that the measure has been calculated for each timeslice, we can connect the geometries for each time slice for the purpose of visualization. 

> Note that the spacetime class' objective is to avoid geometry duplication. But it is necessary for visualization with ggplot2. 

We can cast to an sf object with `as_sf()`.

```{r}
library(ggplot2)

# cast to sf object. Uses merge under the hood
# so if duplicate columns exist, we can specify suffix names
as_sf(bos_gs, suffixes = c("_geo", "_data")) |>
  ggplot(aes(fill = g)) +
  geom_sf(color = NA) +
  facet_wrap("year", nrow = 2) +
  scale_fill_gradient2(
    high = scales::muted("red"),
    low = scales::muted("blue")
    ) + 
  labs(title = "Gun ecometric hotspots",
       subtitle = "2010 - 2019") + 
  theme_void() +
  theme(legend.position = "bottom")
```

> Note that the holes are due to missing data in the original dataset. 

## Point Pattern analysis

There were a few additional functions added which pertain to point pattern analysis. Namely regarding finding centers and creating ellipses based on the centers. These functions supplement gaps in other spatial analysis libraries such as spatstat.

```{r}
library(sf)
library(sfdep)

pnts <- st_sample(guerry, 250)
```

We can calculate the euclidean median center and plot it over all the points. 

```{r}
cent <- euclidean_median(pnts)

plot(pnts)
plot(cent, col = "red", pch = 17, add = TRUE)
```

Additionally, we can identify the standard deviational ellipse for our point set. 

```{r}
ellip_cent <- std_dev_ellipse(pnts)

ellip <- st_ellipse(ellip_cent, 
           ellip_cent$sx, 
           ellip_cent$sy, 
           ellip_cent$theta) 

plot(ellip)
plot(pnts, add = TRUE)
```



## General

- `include_self()`
- `remove_self()`
- Fixes a bug in `local_g_perm()` and `local_gstar_perm()` which would return the G value before being converted into a z-score 




